---
title: "Central Limit Theorem vs Exponential Distribution - A comparitive study"
author: "Lokesh KD"
date: "26 January 2016"
output: html_document
---

### Overview:
    The goal of this project is to investigate, compare and report observations and conclusions  
    made with Central Limit Theorem and Exponential distribution.  
    We shall be working under the following directives:  
    1. Exponential Distribution to be simulated with rexp(n, lambda); lambda - rate parameter.  
    2. Mean and SD is 1/lambda where lambda = 0.2  
    3. Distribution is of 40 exponentials and Simulations is to be carried out are 1000.  

### Simulations:
```{r echo = TRUE}
  # we shall use base graphics to show our plot.
  # first the constants, pre-defined for us.
  sim_lambda <- 0.2
  thr_mean <- 1/sim_lambda  # Theoritical mean.
  sim_dist <- 40
  sim_num <- 1000

  sim_mean <- NULL #Initialization.
  set.seed(17) # A constant to reproduce of our simulations.
  
  # Let us simulate the means for 1000, 40 exponentials each.
  for (i in 1:sim_num) sim_mean <- c(sim_mean, mean(rexp(sim_dist, sim_lambda)))
  
  # Theoritical mean
  thr_mean
  # Simulated mean
  mean(sim_mean)
  
```
    As it can be seen, Theoritical and Simulated mean are pretty close.  
    Let us plot a histogram and visualize.
    
### Sample Mean VS Theoritical Mean:
```{r echo = TRUE}
  # Let us plot the histogram with simulated means
  hist(sim_mean, breaks = 100,
       main = "Histogram of 1000 Simulations, 40 Distributions each",
       xlab = "Mean")
  
  # Let us draw an vertical line with Simulated means.
  abline(v = mean(sim_mean), col = "blue", lwd = 2)
  
  # Now the theoritical mean.
  abline(v= thr_mean, col = "red", lwd = 2)
  
  # Legend
  legend("topright", col = c("blue", "red"),lty = 1,
         legend = c("Simulated Mean", "Theoritical Mean"))
  
```  

### Sample Variance VS Theoritical Variance:  

    Variance is calculated as (mean/sqrt(n))^2 == mean^2/n
    Following the variables defined in our process...
    Theoritical Variance is thr_mean^2/40
    Simulation Variance is (mean(sim_mean))^2/40
    Let us plot the above histogram, this time with variances included.  
    
```{r echo = TRUE}
  # let us calculate the Variance
  
  # Theoritical Variance
  thr_var <- thr_mean^2/sim_dist
  
  # Simulated Variance
  sim_var <- (mean(sim_mean))^2/sim_dist
  
  # Let us plot the histogram with simulated means
  hist(sim_mean, breaks = 100,
       main = "Histogram of 1000 Simulations, 40 Distributions with Variance",
       xlab = "Mean")
  
  # Let us draw an vertical line with Simulated means.
  abline(v = mean(sim_mean), col = "blue", lwd = 2)
  
  # Now the theoritical mean.
  abline(v = thr_mean, col = "red", lwd = 2)
  
  # Theoritical Variance
  abline(v = thr_mean + thr_var, col = "green", lwd = 2)
  abline(v = thr_mean - thr_var, col = "green", lwd = 2)
  
  # Simulated Variance
  abline(v = mean(sim_mean) + sim_var, col = "brown", lwd =2)
  abline(v = mean(sim_mean) - sim_var, col = "brown", lwd =2)
  
  # Legend
  legend("topright", col = c("blue", "red", "green", "brown"),lty = 1,
         legend = c("Simulated Mean", "Theoritical Mean", 
                    "Theoritical Variance", "Simulated Variance"))
  
  
```  

### Distribution:
    With the graphs below, we should clearly see that Simulated Distribution (in blue) is
    pretty close to Normal Distribution(in red). With large data samples, they should
    almost look identical.  
    
```{r echo=TRUE}
  suppressWarnings(library(ggplot2))
  
  g <- ggplot(data.frame(sim_mean), aes(x = sim_mean))
  
  g <- g + labs(title="Distribution(of 40 each) of 1000 Simulations", 
                x="Mean", y="Frequency")

  g <- g + geom_vline(xintercept=thr_mean, color="red", size = 1.0)
                            
  g <- g + geom_vline(xintercept=mean(sim_mean), color="blue", size = 1.0)
                         
  g <- g + stat_function(fun = dnorm, 
                         args = list(mean=thr_mean, sd=sqrt(thr_var)), color = "red")
  
  g <- g + stat_function(fun = dnorm, 
                         args = list(mean=mean(sim_mean), sd=sqrt(sim_var)),color = "blue")
  
  g
  
  
```