---
title: "Tooth Growth Analysis"
author: "Lokesh KD"
date: "27 January 2016"
output: html_document
---

### Overview:
    The goal of this project is to analyze the impact of C vitamin dosage in two forms, i.e
    Orange Juice and a Supplement(tablet?), on the tooth length of Pigs under test.
    
```{r echo = TRUE}
  library(data.table)
  library(lattice)

  # Load the data.
  data("ToothGrowth")
  TG <- data.table(ToothGrowth)
  
  # Let us examine the contents of the data loaded.
  dim(TG)
  str(TG)
  summary(TG)

```  
    So, we have 60 observations with 3 parameters.  
    Two factored observations "OJ" and "VC" as supplements  
    Dose varies with 3 quantities Min = 0.5, Median = 1.0 and Max = 2.0  
    
### Basic Summary:
    Let us examine the mean and standard deviation of the tooth length.  
    Also, plot the impact on Tooth length, based on does, grouped by Supplement.
```{r echo = TRUE}

    mean(TG$len)
    sd(TG$len)
    
    # Plot the graph grouped by Supplement.
    xyplot(dose ~ len | supp, TG, type = "l", layout = c(1,2))

```   

    Per the graph we plotted with the provided sample data observations, it is easier to 
    see that on lower dosages of Vitamin C, i.e 0.5 and 1.0, mean tooth growth is positive.  
    What is interesting is the better impact of Orange Juice on the mean of the tooth growth. 
    
### Hypothesis Testing:
#### Assumptions:
  * Tooth Growth has a positive impact with supplements.  
  * Orange Juice has a better effect on Tooth Growth.
  
#### Hypothesis Testing for OJ vs VC  
```{r echo = TRUE}

    # T-test based on Tooth Growth in length differentiated by supplement.
    t.test(TG$len[TG$supp == "OJ"],TG$len[TG$supp == "VC"], alternative = "greater", 
           paired = FALSE, var.equal = FALSE, conf.level = 0.95)
    
        
```   
    
#### Hypothesis Testing based on the Dosages:  
```{r echo = TRUE}
    
    # T-test based on Tooth Growth in length differentiated by dosage .5 & 1
    t.test(TG$len[TG$dose == .5],TG$len[TG$dose == 1], alternative = "less", 
           paired = FALSE, var.equal = FALSE, conf.level = 0.95)
    
    # T-test based on Tooth Growth in length differentiated by dosage  1 & 2
    t.test(TG$len[TG$dose == 1],TG$len[TG$dose == 2], alternative = "less", 
           paired = FALSE, var.equal = FALSE, conf.level = 0.95)
    
```   

#### Conclusion:  
    1. Based on the T-tests between OJ and VC, we have p-value 0.03, which is higher than  
       0.05 proving Orange Juice has a better impact on Tooth Growth.
    2. Based on T-Tests between dosages 0.5 & 1, 1 & 2 p-values are way below 0.05 proving  
       that higher does of supplements have higer impact.
    
