---
title: "Impact of Weather on Health & Economy in USA - A Data Scientist Approach"
author: "Lokesh KD"
date: "16 January 2016"
output: html_document
---

### Synopsis:
  The goal of this report is to analyze the impact of Weather Events in United States of
  America on Health and Economy across the Country. Data for analysis is taken from NOAA -
  U.S National Oceanic and Atmospheric Administrations during the time period 1950 - 2011.
  
  The Analysis in this report is based on "Fatalities" & "Injuries" for Health, "Crop
  Damage" and "Property Damage" were considered for Economy.
  
  Based on the analysis performed and the results obtained, we could infer that Tornados and
  High Temperatures have great impact on Health while Flood, Drought and Tornados have high
  impact on Economy in United States of America.
   
  Data is obtained from this URL:      
  https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
  
### Data Processing.
  As per the expectations, we start with the raw CSV file.
```{r echo = TRUE}
  #Load the required libraries
  library(plyr)
  library(stats)

  # Check for file existence.
  if (!file.exists("repdata_data_StormData.csv")) {
    if (!file.exists("repdata_data_StormData.csv.bz2")) {
      print("Please download the repdata_data_StormData.csv.bz2 from Assignement page link")
      stop()
    }
    # Unzip the bz2file.
    bunzip2("repdata_data_StormData.csv.bz2")
  }

  # Read the file and load the data.
  WeatherData <- read.csv("repdata_data_StormData.csv", header = TRUE)

  # Examine the data
  dim(WeatherData)
  head(WeatherData)
  
```
  Let us get the subsets to analyze Health and Economic impact.  
  First Health related.
```{r echo=TRUE}
  
  # Get and aggregate subset(s) for Health Parametes impact.
  Fatal <- aggregate(FATALITIES ~ EVTYPE, WeatherData, sum)
  Injury <- aggregate(INJURIES ~ EVTYPE, WeatherData, sum)
```
  Now, onto Economy related...
```{r echo = TRUE}
  # Get subset for Economics with related columns.
  EconomicData <- subset(WeatherData, select = c(EVTYPE, PROPDMG:CROPDMGEXP))
  
  str(EconomicData)
```
  So, we have 19 levels of PROPDMGEXP and 9 levels of CROPDMGEXP.
  We shall compute power of 10 based on the levels we understand
  and multiply and store them in respective DMG variables.
  i.e PROPDMG <- PROPDMG * 10^ (vlaue of PROPDMGEXP)
  i.e CROPDMG <- CROPDMG * 10^ (vlaue of CROPDMGEXP)
  For values we do not understand we substitue 0.
```{r echo = TRUE}
  
  ComputeDamage <- function(Dmg, DmgExp) {
    
    for (i in 1:length(DmgExp)) {
     if (toupper(DmgExp[i]) == "H") # Hunderd
       Dmg[i] <- Dmg[i] * 10 ^ 2
     else if (toupper(DmgExp[i]) == "K") # Kilo
       Dmg[i] <- Dmg[i] * 10 ^ 3
     else if (toupper(DmgExp[i]) == "M") # Million
       Dmg[i] <- Dmg[i] * 10 ^ 6
     else if (toupper(DmgExp[i]) == "B") # Billion
       Dmg[i] <- Dmg[i] * 10 ^ 9
     else if (DmgExp[i] %in% c("" , "-" , "?" , "+")) # Something we do not understand?
       Dmg[i] <- Dmg[i] * 10 ^ 0
     else if (is.numeric(DmgExp[i]))
       Dmg[i] <- Dmg[i] * 10 ^ as.numeric(DmgExp[i])
     else                               
       Dmg[i] <- Dmg[i] * 10 ^ 0
    }
    return (Dmg)
  }
  
  # Call ComputeDamage to re-calculate Property and Crop Damages.
  EconomicData$Pdmg <- ComputeDamage(EconomicData$PROPDMG, EconomicData$PROPDMGEXP)
  EconomicData$Cdmg <- ComputeDamage(EconomicData$CROPDMG, EconomicData$CROPDMGEXP)
  
  # Get the aggregate of the Economic damgages by events
  PropDmg <- aggregate(Pdmg ~ EVTYPE, EconomicData, sum)
  CropDmg <- aggregate(Cdmg ~ EVTYPE, EconomicData, sum)
  
  # Let us look at the new variables created.
  # Check on the correctness of the values in it.
  head(PropDmg)
  head(CropDmg)
```
### Results.
  Here we shall answer the questions as posed in the assignment, now that we have
  loaded and processed the data the way we want to use and represent. we shall 
  barplot the data we have for both the questions.
```{r echo = FALSE}
  cat ("Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?")
```
```{r echo = TRUE}
  # Let us get the top 10 events causing maximum damage to health.
  Fatal10 <- head(arrange(Fatal, desc(FATALITIES)), 10)
  Injury10 <- head(arrange(Injury, desc(INJURIES)), 10)
  
  # Lets look at the top few rows.
  head(Fatal10)
  head(Injury10)
  
  # Health related plots
  par(mfrow=c(1,2), mar=c(8,4,4,2), cex = 0.7)
  
  # Plot the Fatalities
  barplot(Fatal10$FATALITIES, names.arg = Fatal10$EVTYPE, col = "blue", las = 3,
          main = "Top 10 Weather Events resulting in Fatalities",
          ylab = "Number of Fatalities")
  
  # Plor the Injuries
  barplot(Injury10$INJURIES, names.arg = Injury10$EVTYPE, col = "blue", las = 3,
          main = "Top 10 Weather Events resulting in Injuries",
          ylab = "Number of Injuries")
```
```{r echo = FALSE}
  cat("By the histograms, weather event affecting public health most is Tornados for both Fatalities and Injuries")
  cat ("Question 2: Across the United States, which types of events have the greatest economic consequences?")
```
```{r echo = TRUE}
  # Let us get the top 10 events causing maximum damage to property
  # sorted by amount of damage stroed in Pdmg and Cdmg.

  PropDmg10 <- head(arrange(PropDmg, desc(Pdmg)), 10)
  CropDmg10 <- head(arrange(CropDmg, desc(Cdmg)), 10)
  
  # Let us look at the top few lines.
  head(PropDmg10)
  head(CropDmg10)
  
  # Economy related plots
  par(mfrow=c(1,2), mar=c(8,4,4,2), cex = 0.7)
  
  # Plot the Fatalities
  barplot(PropDmg10$Pdmg/10^9, names.arg = PropDmg10$EVTYPE, col = "blue", las = 3,
          main = "Top 10 Weather Events Damaging Property",
          ylab = "Amount of Damage in $ Billions")
  
  # Plor the Injuries
  barplot(CropDmg10$Cdmg/10^9, names.arg = CropDmg10$EVTYPE, col = "blue", las = 3,
          main = "Top 10 Weather Events Damaging Crops",
          ylab = "Amount of Damage in $ Billions")
  
```
```{r echo = FALSE}
  cat("Answer 2: By the histograms above, weather events FLOODS damage Propery the most follwed by Hurricane/Typhoon, Tornado & STROM SURGE. Drought on the other hand damage the Crops most followed by FLOOD, RIVER FLOOD & ICE STORM" )
```